Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array

### 2023-09-18 docs/manuscript/paper.tex
misc paper updates

\subsubsection{Measurement equation}
In general radio pulses from a pulsar are transmitted as amplitude modulations of a radio-frequency carrier wave. They are described by the geometric object $\boldsymbol{p}$ which we identify  as the 4-momentum of the radio photon. The presence of a GW induces a shift in the temporal component of the covariant 4-momentum, while the photon travels from the emitter to the observer, i.e. $\Delta p_t = p_t|_{\rm observer} - p_t|_{\rm emitter} $. One obtains \citep[e.g.][]{Maggiore}
\begin{equation}
 \Delta p_t =  \frac{\pi f_{\rm p} h_{ij} (t; \boldsymbol{x}= 0)q^i q^j }{1 + \boldsymbol{n}\cdot \boldsymbol{q} }  \left[1 -e^{i \Omega (1 + \boldsymbol{n}\cdot \boldsymbol{q})  d}\right] \ ,
	\label{eq:momentum_shift}
\end{equation}
where $f_{\rm p}$ is the pulse frequency measured in the momentarily comoving reference frame of an observer. We write $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar. We take the pulsar location to be constant i.e.  neither $\boldsymbol{q}$ nor $d$ are functions of time. In practice the pulsar locations vary with respect to the Earth but are constant with respect to the SSB. The barycentering correction is typically applied when generating TOAs, e.g. with {\sc tempo2} \citep{tempo2} and related timing software, and is inherited by the frequency time series. Some pulsars, including some PTA pulsars, do have non-negligible proper motions $\mathcal{O} (100)$ km s$^{-1}$ after the barycentering corrections have been applied \citep[e.g.][]{10.1093/mnras/sty3390}, but we do not consider this effect in this paper. Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\boldsymbol{u}$ is given by the coordinate-independent expression $p_{\alpha} u^{\alpha}$. After barycentering, one has $u^{\alpha} =(1,0,0,0)$ for both the emitter and the observer to leading order in the respective momentarily comoving reference frames,
\begin{equation}
	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm observer} = (1,0,0,0) \ ,
\end{equation}
Hence Equation \eqref{eq:momentum_shift} can be written for the $n$-th pulsar as
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d^{(n)}) g^{(n)}(t) +  \varepsilon^{(n)}\ .
	\label{eq:measurement}
\end{equation}
where $d^{(n)}$ labels the distance to the $n$-th pulsar, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies: 
\begin{align}
	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'}(t - t') \ .	\label{eq:vareps}
\end{align}
where $\sigma_{\rm m}$ is the covariance of the measurement noise and is shared between all pulsars. The measurement function $g^{(n)}(t)$ is
\begin{equation}
	g^{(n)}(t) = 1 - \frac{h_{ij} (t; \boldsymbol{x}= 0)[q^{(n)}]^i [q^{(n)}]^j}{2[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] }  \left \{1 -e^{i \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}}\right \} \ ,
	\label{eq:g_func}
\end{equation}
and $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$. It is also instructive to express Equation \eqref{eq:g_func} in a trigonometric form as,
\begin{align}
	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
	&- \cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \Big ] \ .
	\label{eq:g_func_trig}
\end{align}
Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}  define a non-linear measurement equation that relates the intrinsic pulsar spin frequency to the pulse frequency measured by an observer on Earth.

@@

	\includegraphics[width=\columnwidth]{images/example_residuals_plot}

@@

\section{Kalman recursion equations} \label{sec:kalman}
The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k$, which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$, via a linear transformation:
\begin{equation}
	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \ ,\label{eq:kalman1}
\end{equation}
where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step. The Kalman filter takes the underlying states to evolve according to
\begin{equation}
	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \ , \label{eq:kalman2}
\end{equation}
where $\boldsymbol{F}$ is the system dynamics matrix, $\boldsymbol{G}$ the control matrix. $\boldsymbol{u}$ the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise with covariance $\boldsymbol{Q}_k$. \newline

The Kalman filter is a recursive estimator with two distinct stages; a ``predict" stage where the state at timestep $k+1$ is predicted based on the state at timestep $k$, and an ``update" stage where the predicted state estimate is refined based on the measurement at that time-step. The predict stage updates the estimate of the state $\hat{\boldsymbol{X}}_{k|k-1}$ and the associated covariance $\hat{\boldsymbol{P}}_{k|k-1}$where the subscript notation ${k|k-1}$ denotes an estimate at discrete step $k$, given measurements from step $k-1$ and earlier. The predict step proceeds as,
\begin{equation}
\hat{\boldsymbol{X}}_{k|k-1} =  \boldsymbol{F}_k \hat{\boldsymbol{X}}_{k-1|k-1} + \boldsymbol{G}_k \boldsymbol{u}_k
\end{equation}
\begin{equation}
	\hat{\boldsymbol{P}}_{k|k-1} =  \boldsymbol{F}_k \hat{\boldsymbol{P}}_{k-1|k-1} \boldsymbol{F}_k^\intercal + \boldsymbol{Q}_k 
\end{equation}
The update step then uses the measurement $\boldsymbol{Y}_k$ to update $\hat{\boldsymbol{X}}_{k|k}$  and $\hat{\boldsymbol{P}}_{k|k}$ as,
\begin{equation}
	\boldsymbol{\epsilon}_{k} = \boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k-1}
\end{equation} 
\begin{equation}
	\boldsymbol{S}_k = \boldsymbol{H}_k \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal + \boldsymbol{R}_k
\end{equation}
\begin{equation}
	\boldsymbol{K}_k = \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal \boldsymbol{S}_k^{-1} \label{eq:kalman gain}
\end{equation}
\begin{equation}
	\hat{\boldsymbol{X}}_{k|k} =\hat{\boldsymbol{X}}_{k|k-1} +\boldsymbol{K}_k  \boldsymbol{\epsilon}_{k}  \label{eq:kalmangainupdate}
\end{equation}
\begin{equation}
		\hat{\boldsymbol{P}}_{k|k} = \left( \boldsymbol{I} - \boldsymbol{K}_k \boldsymbol{H}_k \right) 	\hat{\boldsymbol{P}}_{k|k-1}
\end{equation}
Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which yields the minimal mean square error in the measurement post fit residual, $\boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k}$. \newline

We map our continuous-time state-space model described in Section \ref{sec:model} onto the discrete-time Kalman filter structure of Equations  
\eqref{eq:kalman1}, \eqref{eq:kalman2} as follows. We identify $\boldsymbol{X}$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e 
\begin{equation}
	\boldsymbol{X} = \left(f_{\rm p}^{(1)}, f_{\rm p}^{(2)}, ..., f_{\rm p}^{(N)}\right) \ .
\end{equation}
Similarly,  
\begin{equation}
\boldsymbol{Y} = \left(f_{\rm m}^{(1)}, f_{\rm m}^{(2)}, ..., f_{\rm m}^{(N)} \right)
\end{equation}
The states evolve according the continuous dynamical equation
\begin{equation}
	d \boldsymbol{X} = \boldsymbol{A} \boldsymbol{X} dt + \boldsymbol{C}(t) dt + \boldsymbol{\Sigma} d \boldsymbol{B}(t) \label{eq:kalmn2}
\end{equation}
where $\boldsymbol{A} = \text{diag} \left(\gamma^{(1)}, \gamma^{(2)}, ..., \gamma^{(N)}\right)$ is the state matrix, $\boldsymbol{C} = \left(C^{(1)}, C^{(2)}, ..., C^{(N)}\right)$ the input or control matrix where $C^{(i)} =\gamma^{(i)} \left(f_{\rm em}^{(i)} (0) + \dot{f}_{\rm em}(0)^{(i)} t \right) +  \dot{f}_{\rm em}(0)^{(i)}$ and $d\boldsymbol{B}(t)$ denotes a Wiener process with covariance matrix $\boldsymbol{\Sigma} = \text{diag} \left(\sigma^{(1)}, \sigma^{(2)}, ..., \sigma^{(N)}\right)$. Equation \ref{eq:kalmn2} is a Langevin equation (equivalently an Ornstein-Uhlenbeck process) which has a general solution given by \citep{gardiner2009stochastic},
\begin{equation}
	\boldsymbol{X}(t) = e^{\boldsymbol{A} t} \boldsymbol{X}(0) + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{C}(t') dt' + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{\Sigma} d\boldsymbol{B}(t') \label{eq:gardenier}
\end{equation} 
From Equation \eqref{eq:gardenier} we can construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ as Equation \eqref{eq:kalman2} where
\begin{equation}
\boldsymbol{F}_k = e^{\boldsymbol{A}\left( t_{k+1} - t_k \right)} 
\end{equation}
\begin{equation}
	\boldsymbol{G}_k = \int_{t_i}^{t_{i+1}}  e^{\boldsymbol{A}\left( t_{k+1} - t' \right)}  \boldsymbol{C}(t') dt' 
\end{equation}
\begin{equation}
	\boldsymbol{w}_k = \int_{t_k}^{t_{k+1}} e^{\boldsymbol{A}\left( t_{k+1} - t' \right)} \boldsymbol{\Sigma} d \boldsymbol{B}(t')
	\end{equation}
For our system, the state, control and process noise covariance matrices then take the form:
\begin{equation}
	\boldsymbol{F}_k = 	\text{diag}\left(e^{- \gamma^{(1)} \Delta t},e^{- \gamma^{(2)} \Delta t},...,e^{- \gamma^{(N)} \Delta t} \right)
\end{equation}
\begin{equation}
\boldsymbol{G}_k	= \left(G^{(1)}_k, G^{(2)}_k,...,G^{(N)}_k \right)
\end{equation}
where
\begin{equation}
	G_k^{(n)} =    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left[\Delta t + t_k \right] - e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right]
\end{equation}
and
\begin{equation}
	\boldsymbol{Q}_k \boldsymbol{\delta}_{kj}= \langle \boldsymbol{\eta}_k \boldsymbol{\eta}_j^\intercal \rangle = \text{diag} \left(Q^{(1), Q^{(2)}},...,Q^{(N)}\right) 
\end{equation}
for 
\begin{equation}
	Q^{(n)} = \frac{- [\sigma^{n}]^2}{2 \gamma^{(n)}} \left( e^{-2 \gamma^{(n)} \Delta t} -1\right)
\end{equation}
where $\Delta t$ is the constant time sampling interval $=t_{k+1} -t_{k}$. \newline

The remaining component matrices of the Kalman filter are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$. These are defined straightforwardly from Equation \eqref{eq:measurement}. Specifically, 
$\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \eqref{eq:g_func}. $\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$.

>>>>

\subsubsection{Measurement equation}
In general radio pulses from a pulsar are transmitted as amplitude modulations of a radio-frequency carrier wave. They are described by the geometric object $\boldsymbol{p}$ which we identify  as the 4-momentum of the radio pulse. The presence of a GW induces a shift in the temporal component of the covariant 4-momentum, while the photon travels from the emitter to the observer, i.e. $\Delta p_t = p_t|_{\rm observer} - p_t|_{\rm emitter} $. One obtains \citep[e.g.][]{Maggiore}
\begin{equation}
 \Delta p_t =  \frac{\pi f_{\rm p} h_{ij} (t; \boldsymbol{x}= 0)q^i q^j }{1 + \boldsymbol{n}\cdot \boldsymbol{q} }  \left[1 -e^{i \Omega (1 + \boldsymbol{n}\cdot \boldsymbol{q})  d}\right] \ ,
	\label{eq:momentum_shift}
\end{equation}
where $f_{\rm p}$ is the pulse frequency measured in the momentarily comoving reference frame of an observer. We write $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar. We take the pulsar location to be constant i.e.  neither $\boldsymbol{q}$ nor $d$ are functions of time. In practice the pulsar locations vary with respect to the Earth but are constant with respect to the SSB. The barycentering correction is typically applied when generating TOAs, e.g. with {\sc tempo2} \citep{tempo2} and related timing software, and is inherited by the frequency time series. Some pulsars, including some PTA pulsars, do have non-negligible proper motions $\mathcal{O} (100)$ km s$^{-1}$ after the barycentering corrections have been applied \citep[e.g.][]{10.1093/mnras/sty3390}, but we do not consider this effect in this paper. Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\boldsymbol{u}$ is given by the coordinate-independent expression $p_{\alpha} u^{\alpha}$. After barycentering, one has $u^{\alpha} =(1,0,0,0)$ for both the emitter and the observer to leading order in the respective momentarily comoving reference frames,
\begin{equation}
	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm observer} = (1,0,0,0) \ ,
\end{equation}
Hence Equation \eqref{eq:momentum_shift} can be written for the $n$-th pulsar as
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d^{(n)}) g^{(n)}(t) +  \varepsilon^{(n)}\ .
	\label{eq:measurement}
\end{equation}
where $d^{(n)}$ labels the distance to the $n$-th pulsar, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies: 
\begin{align}
	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'}(t - t') \ .	\label{eq:vareps}
\end{align}
where $\sigma_{\rm m}$ is the covariance of the measurement noise and is shared between all pulsars. The measurement function $g^{(n)}(t)$ is
\begin{equation}
	g^{(n)}(t) = 1 - \frac{h_{ij} (t; \boldsymbol{x}= 0)[q^{(n)}]^i [q^{(n)}]^j}{2[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] }  \left \{1 -e^{i \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}}\right \} \ ,
	\label{eq:g_func}
\end{equation}
and $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$. It is also instructive to express Equation \eqref{eq:g_func} in a trigonometric form as,
\begin{align}
	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
	&- \cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \Big ] \ .
	\label{eq:g_func_trig}
\end{align}
Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}  define a non-linear measurement equation that relates the intrinsic pulsar spin frequency to the pulse frequency measured by an observer on Earth.

@@

	\includegraphics[width=\columnwidth]{images/example_residuals_plot2}

@@

\section{Kalman recursion equations} \label{sec:kalman}
The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k$, which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$, via a linear transformation:
\begin{equation}
	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \ ,\label{eq:kalman1}
\end{equation}
where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance, and the subscript $k$ labels the time-step. The Kalman filter takes the underlying states to evolve according to
\begin{equation}
	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \ , \label{eq:kalman2}
\end{equation}
where $\boldsymbol{F}$ is the system dynamics matrix, $\boldsymbol{G}$ the control matrix. $\boldsymbol{u}$ the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise, $\mathcal{N} \sim (0,\boldsymbol{Q}_k)$. \newline

The Kalman filter is a recursive estimator with two distinct stages; a ``predict" stage and an ``update" stage. The predict stage predicts the estimate of the state at discrete step $k$, given the state estimates from step $k-1$. Specifically, the predict step proceeds as,
\begin{equation}
	\hat{\boldsymbol{X}}_{k|k-1} =  \boldsymbol{F}_k \hat{\boldsymbol{X}}_{k-1|k-1} + \boldsymbol{G}_k \boldsymbol{u}_k
\end{equation}
\begin{equation}
	\hat{\boldsymbol{P}}_{k|k-1} =  \boldsymbol{F}_k \hat{\boldsymbol{P}}_{k-1|k-1} \boldsymbol{F}_k^\intercal + \boldsymbol{Q}_k 
\end{equation}
Note that the predict stage is independent of the measurements. The measurement $\boldsymbol{Y}_k$ is included to update the prediction during the update stage as follows:
\begin{equation}
	\boldsymbol{\epsilon}_{k} = \boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k-1}
\end{equation} 
\begin{equation}
	\boldsymbol{S}_k = \boldsymbol{H}_k \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal + \boldsymbol{R}_k
\end{equation}
\begin{equation}
	\boldsymbol{K}_k = \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal \boldsymbol{S}_k^{-1} \label{eq:kalman gain}
\end{equation}
\begin{equation}
	\hat{\boldsymbol{X}}_{k|k} =\hat{\boldsymbol{X}}_{k|k-1} +\boldsymbol{K}_k  \boldsymbol{\epsilon}_{k}  \label{eq:kalmangainupdate}
\end{equation}
\begin{equation}
		\hat{\boldsymbol{P}}_{k|k} = \left( \boldsymbol{I} - \boldsymbol{K}_k \boldsymbol{H}_k \right) 	\hat{\boldsymbol{P}}_{k|k-1}
\end{equation}
Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which is defined so as to minimise the mean squared error in the state estimate, i.e.  $\boldsymbol{K}_k = \text{argmin} \left[ \boldsymbol{E}(\boldsymbol{X}_k - \hat{\boldsymbol{X}}_k) \right ]$. For a full review of the Kalman filter, including its derivation, we refer the reader to \cite{Gelb:1974} and \cite{zarchan2000fundamentals}. \newline

We map our continuous-time state-space model described in Section \ref{sec:model} onto the discrete-time Kalman filter structure of Equations  
\eqref{eq:kalman1}, \eqref{eq:kalman2} as follows. We identify $\boldsymbol{X}$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e 
\begin{equation}
	\boldsymbol{X} = \left(f_{\rm p}^{(1)}, f_{\rm p}^{(2)}, ..., f_{\rm p}^{(N)}\right) \ .
\end{equation}
Similarly,  
\begin{equation}
\boldsymbol{Y} = \left(f_{\rm m}^{(1)}, f_{\rm m}^{(2)}, ..., f_{\rm m}^{(N)} \right)
\end{equation}
The states evolve according the continuous dynamical equation (c.f. Equation \eqref{eq:frequency_evolution})
\begin{equation}
	d \boldsymbol{X} = \boldsymbol{A} \boldsymbol{X} dt + \boldsymbol{C}(t) dt + \boldsymbol{\Sigma} d \boldsymbol{B}(t) \label{eq:kalmn2}
\end{equation}
where $\boldsymbol{A}$ is a diagonal $N \times N$ square matrix
\begin{equation}
\boldsymbol{A} = \text{diag} \left(\gamma^{(1)}, \gamma^{(2)}, ..., \gamma^{(N)}\right) \ ,
\end{equation}
and $\boldsymbol{C}(t)$ a time dependent vector with $n$-th component
\begin{equation}
 C^{(n)} =\gamma^{(n)} \left(f_{\rm em}^{(n)} (t_1) + \dot{f}_{\rm em}(t_1)^{(n)} t \right) +  \dot{f}_{\rm em}(t_1)^{(n)} \ .
\end{equation}
The $N \times N$ square matrix $\boldsymbol{\Sigma}$  governs the magnitude of the increments of Brownian motion (Wiener process) $d\boldsymbol{B}(t)$, where
\begin{equation}
\boldsymbol{\Sigma} = \text{diag} \left(\sigma^{(1)}, \sigma^{(2)}, ..., \sigma^{(N)}\right)
\end{equation}
\newline

The evolution of the state is an Ornstein-Uhlenbeck process, described by a Langevin equation, Equation \eqref{eq:kalmn2}, which has a general solution given by \citep{gardiner2009stochastic},
\begin{equation}
	\boldsymbol{X}(t) = e^{\boldsymbol{A} t} \boldsymbol{X}(0) + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{C}(t') dt' + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{\Sigma} d\boldsymbol{B}(t') \label{eq:gardenier}
\end{equation} 
From Equation \eqref{eq:gardenier} we can construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ in the form of Equation \eqref{eq:kalman2} where
\begin{equation}
\boldsymbol{F}_k = e^{\boldsymbol{A}\left( t_{k+1} - t_k \right)} 
\end{equation}
\begin{equation}
	\boldsymbol{G}_k = \int_{t_i}^{t_{i+1}}  e^{\boldsymbol{A}\left( t_{k+1} - t' \right)}  \boldsymbol{C}(t') dt' 
\end{equation}
\begin{equation}
	\boldsymbol{w}_k = \int_{t_k}^{t_{k+1}} e^{\boldsymbol{A}\left( t_{k+1} - t' \right)} \boldsymbol{\Sigma} d \boldsymbol{B}(t')
	\end{equation}
For our system, the state, control and process noise covariance matrices then take the form:
\begin{equation}
	\boldsymbol{F}_k = 	\text{diag}\left(e^{- \gamma^{(1)} \Delta t},e^{- \gamma^{(2)} \Delta t},...,e^{- \gamma^{(N)} \Delta t} \right)
\end{equation}
\begin{equation}
\boldsymbol{G}_k	= \left(G^{(1)}_k, G^{(2)}_k,...,G^{(N)}_k \right)
\end{equation}
with
\begin{align}
	G_k^{(n)} =&    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left[\Delta t + t_k \right] \nonumber \\ 
	&- e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right]
\end{align}
and
\begin{equation}
	\boldsymbol{Q}_k \boldsymbol{\delta}_{kj}= \langle \boldsymbol{\eta}_k \boldsymbol{\eta}_j^\intercal \rangle = \text{diag} \left(Q^{(1)}, Q^{(2)},...,Q^{(N)}\right) 
\end{equation}
for 
\begin{equation}
	Q^{(n)} = \frac{- [\sigma^{n}]^2}{2 \gamma^{(n)}} \left( e^{-2 \gamma^{(n)} \Delta t} -1\right)
\end{equation}
where $\Delta t$ is the constant time sampling interval $=t_{k+1} -t_{k}$. \newline

The remaining component matrices of the Kalman filter that we have not yet specified are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$. These are defined straightforwardly from Equation \eqref{eq:measurement}. Specifically, 
$\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \eqref{eq:g_func}. The measurement covariance $\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$.
